Double White Dwarf Mergers with CASTRO\\ I. Methodology and Code Verification

### 2015-01-28 papers/wdmerger_I/paper.tex
Rewrite of the gravity source term section.

The effect of gravity on the hydrodynamical evolution is typically
incorporated by the use of a source term for the momentum and energy
equations. The integral form of the momentum source term is
\begin{equation}
  \int \rho \mathbf{g}\, dV
\end{equation}
and for the energy source term it is
\begin{equation}
  \int \rho \mathbf{u}\cdot\mathbf{g}\, dV \label{eq:cell_center_gravity_source}.
\end{equation}
In most hydrodynamics codes these are discretized as $\rho\,
\mathbf{g}\, \Delta V$ and $\rho\, \mathbf{u}\,\cdot\mathbf{g}\,
\Delta V$, respectively, where $\Delta V$ is the zone volume and
$\rho$, $\mathbf{u}$, and $\mathbf{g}$ are evaluated at the zone
center. As observed by \cite{arepo}, the essential problem with this
approach is that fluid movements in a grid code are calculated using
fluxes at zone boundaries, while the standard gravity source term is
an estimate of the flux using the zone center quantities. This can
lead to a striking mismatch when the gradient across a zone is
large. We have found that this is especially important at the edge of
a star, where the large density gradient means that the momentum and
energy fluxes can transition sharply even across half of a zone
width. Another problem is that in a merger calculation in which two
stars are moving on a grid, large quantities of material are being
advected, so even a slight error in the calculation for the
high-density material can manifest itself as a long-term error in
energy conservation. This becomes apparent in Section
\ref{sec:kepler}. So while a closer look at this may be unwarranted
for situations like a stationary single star central explosion
calculation, a detailed discussion is certainly called for in the case
of an orbiting binary system where getting the gravity right is
absolutely necessary for fully trusting the long-term evolution of the
system.

\citet[Chapter 4]{shu:1992} observes that it is possible to describe the source term for the momentum equation by taking the divergence of a gravitational stress tensor,
\begin{equation}
  G_{ij} = -\frac{1}{4\pi G}\left(g_i g_j - \frac{1}{2}|\mathbf{g}|^2\delta_{ij}\right).
\end{equation}
The momentum equations are then written explicitly in conservative
form when discretized: just as with the momentum stress tensor,
divergences are calculated using discretized fluxes at zone
boundaries, and the flux at any zone boundary is added to one cell and
subtracted from another, so that the total momentum stays constant to
within numerical roundoff error (with the possible exception of domain
boundaries). This method works because, from the Poisson equation,
$\rho$ can be written as proportional to the divergence of
$\mathbf{g}$. We note that in order for the method to be reliable, the
expression for all components of $\mathbf{g}$ must be correctly
calculated from $\phi$ on zone boundaries so that the total divergence
does indeed recover the density. This is straightforward for the
component normal to the zone interface, but requires averaging over
zones in the perpendicular direction to obtain the transverse
components.

Dealing with the energy equation is more difficult. The central
challenge is to write down a form of the discretized energy equation
that explicitly conserves total energy when coupled to gravity. When
gravity is included, the total energy is
\begin{equation}
  \rho E_{\text{tot}} = E_G - \frac{1}{2}\rho\phi = \frac{1}{2}\rho \mathbf{u}^2 + \rho e - \frac{1}{2}\rho\phi,
\end{equation}
where $E_G$ is the gas energy from the pure hydrodynamics equation,
and the factor of $1/2$ in the gravitational energy term is necessary
for simulations with self-gravity to prevent double-counting of
interactions. Recently, three different approaches have been proposed
for achieving this. We will briefly examine these and then describe
the approach that makes the most sense for us.

\citet{marcello:2012} use the mass continuity equation to rewrite the
standard source term from Equation
\ref{eq:cell_center_gravity_source}. When this is done, the resulting
evolution equation for the total energy is:
\begin{equation}
  \frac{\partial (\rho E_{\text{tot}})}{\partial t} = -\bm{\nabla}\cdot\left(\rho \mathbf{u} E_{\text{tot}} + p\mathbf{u} - \frac{1}{2}\rho\mathbf{u}\phi\right) + \frac{1}{2}\phi\frac{\partial \rho}{\partial t} - \frac{1}{2}\rho\frac{\partial \phi}{\partial t}.
\end{equation}
In practice, we would still actually evolve the total gas energy
$E_G$, and so when discretized to second order in time from time level
$n$ to time level $n+1$ the gas energy equation would look like
\begin{align}
  (\rho E_G)^{n+1} &= (\rho E_G)^{n} + \left(\frac{1}{2} (\rho \phi)^{n+1} - \frac{1}{2} (\rho \phi)^{n}\right) \notag \\
   &- \Delta t\, \bm{\nabla}\cdot\left(\rho \mathbf{u} E_{\text{tot}} + p\mathbf{u} - \frac{1}{2}\rho\mathbf{u}\phi\right) \notag \\
   &+ \frac{\Delta t}{2}\left(\phi \frac{\partial \rho}{\partial t}\right)^{n+1/2} - \frac{\Delta t}{2} \left(\rho \frac{\partial \phi}{\partial t}\right)^{n+1/2}.
\end{align}
It is not immediately obvious that this is explicitly conservative for
the total energy, due to the last two terms on the right-hand side,
but written in discrete form the potential is linked to the density in
such a way that there is indeed a perfect cancellation of terms
between adjacent cells (see Equation 88 in
\citeauthor{marcello:2012}). Doing it this way requires a prescription
for calculating $\phi^{n+1/2}$, the potential at the time halfway
between the new and old state, as well as $\dot{\phi}^{n+1/2}$, the
time derivative of the potential at that time. To second order in
time, these can be calculated as $(\phi^{n+1} + \phi^{n})/2$ and
$(\phi^{n+1}-\phi^{n})/\Delta t$ respectively. So, in practice what we
do is to first solve the Poisson equation using $\rho^n$ at the
beginning of a time step; then, evolve the pure hydrodynamics
equations without the gravitational source to obtain $\rho^{n+1}$;
then, solve the Poisson equation for $\phi^{n+1}$; and then to finally
calculate the gravitational source terms and add them in. An
alternative approach is to use the predictor-corrector scheme that
CASTRO features, where the time level $n$ source term is included in
the hydrodynamics update and then left out of the final update. In
either case, the time level $n$ calculation of $\mathbf{g}$ is
included the PPM reconstruction of the edge states.

\citet{jiang:2013} start from the motivation of extending to the
energy equation the flux-based formalism that allows explicit
conservation for the momentum equation. They determine that the energy
is perfectly conserved if we define a gravitational flux
$\mathbf{F_g}$ and take its divergence for the energy equation:
\begin{align}
  \mathbf{F_g} &= \frac{1}{8\pi G}\left(\phi\bm{\nabla}\dot{\phi} - \dot{\phi}\bm{\nabla}\phi\right) + \rho\mathbf{u}\phi \label{eq:jiang_flux}\\
  \frac{\partial (\rho E_G)}{\partial t} &= -\bm{\nabla} \cdot \left(\rho\mathbf{u}E_G + p\mathbf{u} + \mathbf{F_g}\right)
\end{align}
This is analytically equivalent to the approach by
\citet{marcello:2012}, which can be seen by taking the divergence of
the flux and using the Poisson equation. It would be justified to use
this form simply because it is cleaner to write down, but there is a
more important benefit to using an explicit flux based formula: in
AMR, conservation between levels can only happen if all fluid
movements are written down using fluxes, because it is those fluxes
that are adjusted at coarse-fine interfaces to ensure conservation. In
this form we can simply add the gravitational source term to the
hydrodynamical fluxes and use the existing AMR algorithm to guarantee
conservation between levels. We note that \citeauthor{jiang:2013} say
that this method requires the solution of an additional Poisson
equation to obtain the intermediate time gravitational potential, but
as we pointed out above, this can be done to second order simply by
taking a centered difference using the existing Poisson solves, so no
extra Poisson solve is necessary.

This flux-based formalism has a significant drawback for our problem:
unlike \citeauthor{jiang:2013}, we use open hydrodynamic boundary
conditions, not periodic boundary conditions. Mass, momentum and
energy can be lost (or gained) across the domain boundary. This is not
a serious source of error in the standard cell-centered
discretization, because the source term is proportional to $\rho$ and
the density near the domain boundary is generally filled with a very
low density ambient medium, so the size of the source term is always
small. However, the gravitational flux in Equation \ref{eq:jiang_flux}
has terms that are not explicitly proportional to
density. Consequently, the flux lost across the domain boundaries can
be significant (though this effect diminishes with increasing
resolution). A related problem is that at the edge of a star, the
gravitational flux can be so large that the gas energy becomes
negative after an update. This effect is greatly diminished if the
flux is instead proportional to density, as in the standard
method. This is not a major problem for ideal or polytropic gases in
pure hydrodynamics problems, but in our simulations we want the
internal energy (and thus the temperature) to always be reliable so
that nuclear reactions occur at the correct rates. This is especially
sensitive at the edges of the stars, where the burning is expected to
occur in a violent merger scneario.

A possible compromise approach is discussed in \cite{arepo}. We start
with the integral formulation of the energy source term (Equation
\ref{eq:cell_center_gravity_source}), but instead of discretizing at
the cell center, we use Gauss' theorem to evaluate the source term at
the zone boundaries (see Equations 91--96 of that paper; we exclude
the term related to the motion of the zones since our grids are
fixed). This results in a gravitational update for zone $i$ written as
\begin{equation}
  \Delta E^{\text{grav}}_{i} = -\frac{1}{2}\sum_{j} \Delta m_{ij}\, \mathbf{r}_{ij} \cdot \bm{\nabla}\phi_i \approx -\frac{1}{2}\sum_{j} \Delta m_{ij}\left(\phi_i - \phi_j\right),
\end{equation}
where the sum is over all $j$ zones that share a boundary with zone
$i$. Here $\mathbf{r}_{ij}$ is the radius vector joining zones $i$ and
$j$ and $\Delta m_{ij}$ is the mass exchanged between zones $i$ and
$j$ in the hydrodynamics update, and is equal to $\Delta t A_{ij}
F_\rho^{ij}$, where $A_{ij}$ is the area of the zone boundary and
$F_{\rho}^{ij}$ is the mass flux from the mass continuity equation. It
is straightforward to see that the second version is explicitly
conservative. The benefit of this form is that the mass exchanged
between zones near the domain boundary will always be small in
absolute terms since these zones are at very low density. Thus the
energy lost through the boundary will be of comparable size to the
standard method, but the gravity update should be more accurate since
the change in gravitational energy now tracks the explicit motion of
mass on the grid.

% Now rewrite this so that it is in flux form

@@

Since the results of a merger simulation depend strongly on gravity,
it is important to check whether proper numerical convergence is
achieved for the Poisson solver. To do so, we created a simple test
that inserts a sphere of radius $R$ and uniform mass density $\rho$
onto our grid, and used CASTRO to calculate the gravitational
potential $\phi$ of this setup. We ensure that $R$ is an integer
multiple of the grid spacing, and the center of the sphere is at the
origin. The problem domain for our simulations is $[-1.6, 1.6]^3$, and
we take $R = 0.5$ and $\rho = 10^3$. The zones with $r > R$ are filled
with an ambient material of very low density. We run this problem at
multiple resolutions corresponding to jumps by a factor of two. For
comparison, at each grid point we evaluate the analytical potential of
a uniform sphere, which can be easily determined using Gauss' law:
\begin{equation}
  \phi_{\text{sphere}}(r) = \frac{GM}{r} \times \begin{cases} (3R^2 - r^2)/(2 r^2) & r \leq R \\ 1 & r > R \end{cases},\label{eq:sphere-analytical}
\end{equation}
where $M = 4\pi R^3 / 3$ is the mass of the sphere. Defining the $L^2$
norm of a field $f$ as
\begin{equation}
  \| f \|_2 = \left(\sum_{i,j,k} dx\, dy\, dz\, f_{ijk}^2\right)^{1/2},
\end{equation}
we measure the numerical error by calculating the $L^2$ norm of the error and normalizing it by the $L^2$ norm of the analytical solution:
\begin{equation}
  \text{Error} = \frac{\|\phi - \phi_{\text{sphere}}\|_2}{\|\phi_{\text{sphere}}\|_2}.
\end{equation}

@@

The only way to fully eliminate this effect is to test a problem that
does not change with resolution. The obvious companion is a cube of
uniform density $\rho$, where now $R$ is half of the side length of
the cube. At each resolution we use the same $R$ as for the sphere,
which ensures that the cube always fills exactly the same fraction of
the domain and thus has the same mass, so the only improvement comes
from better sampling at higher resolution. The potential for this
object has been worked out analytically by \citet{waldvogel:1976} (see
also a similar result in \citealt{hummer:1996}). The potential is in
Equation 15 of that paper, though the last term is missing a factor of
$1/2$, which destroys the symmetry. Inserting this missing factor and
performing a simple coordinate transformation so that the center of
the cube is at the origin, the potential is
\begin{align}
  \phi_{\text{cube}}(x,y,z) &= G\rho\sum_{i,j,k=0}^1\left[x_i y_j\, \text{tanh}^{-1}\left(\frac{z_k}{r_{ijk}}\right)\right. \notag \\
  &+ \left. y_j z_k\, \text{tanh}^{-1}\left(\frac{x_i}{r_{ijk}}\right) + z_k x_i\, \text{tanh}^{-1}\left(\frac{y_j}{r_{ijk}}\right) \right.\notag \\
  &\left. - \frac{x_i^2}{2}\,\text{tan}^{-1}\left(\frac{y_j z_k}{x_i r_{ijk}}\right) - \frac{y_j^2}{2}\,\text{tan}^{-1}\left(\frac{z_k x_i}{y_j r_{ijk}}\right) \right. \notag \\
  &- \left. \frac{z_k^2}{2}\,\text{tan}^{-1}\left(\frac{x_i y_j}{z_k r_{ijk}}\right)\right]
\end{align}
where $x_0 = R + x$, $x_1 = R - x$, $y_0 = R + y$, $y_1 = R-y$, $z_0 =
R+z$, $z_1 = R-z$, and $r_{ijk} = \sqrt{x_i^2 + y_j^2 + z_k^2}$. In
\texttt{FORTRAN} and \texttt{C}, the inverse hyperbolic tangent is
\texttt{atanh} and the inverse tangent is \texttt{atan} (\textit{not}
\texttt{atan2}). This formula is valid both inside and outside the
cube. The normalized $L^2$ error norm for this problem is also shown
in Figure \ref{fig:gravity_convergence}, and indeed we see
near-perfect second-order scaling.

@@

The tracing of gravity can be controlled in the public version of Castro
with the parmeter {\tt castro.ppm\_trace\_grav}.

\subsection{Gravity Source Correction}

We also modify slightly the correction done to the gravitational
source terms after integrating the system in time.  In the original
Castro paper (Eqs. 19 and 20), the state after the conservative update
due to fluxes and external sources (other than gravity) was
represented with the superscript `$(2,\star)$'.  The gravitational
sources were then corrected, making them centered in time, yielding
the final state at the end of the hydrodynamics step, indicated with
the superscript `$(2)$'.  This appeared as:
\begin{eqnarray}
(\rho {\bf U})^{(2)} &=& (\rho {\bf U})^{(2,\star)} +
   \frac{\Delta t}{2} \left [ (\rho {\bf g})^{(2,\star)} - 
                              (\rho {\bf g})^{(1)} \right ]\\
(\rho E)^{(2)} &=& (\rho E)^{(2,\star)} +
   \frac{\Delta t}{2} \left [ (\rho {\bf U\cdot g})^{(2,\star)} - 
                              (\rho {\bf U\cdot g})^{(1)} \right ]
\end{eqnarray}
Here we make the subtle change of using the corrected ${\bf u}$ in the
correction to the energy.  So our new energy correction equation appears as:
\begin{equation}
(\rho E)^{(2)} = (\rho E)^{(2,\star)} +
   \frac{\Delta t}{2} \left [ (\rho {\bf U\cdot g})^{(2)} - 
                              (\rho {\bf U\cdot g})^{(1)} \right ]
\end{equation}
Both of these are second-order accurate, but we've found that the
latter is slightly better for conserving energy.  This change is now
the default in the public version of Castro (it can be controlled by
{\tt castro.grav\_source\_type}.

>>>>

The effect of gravity on the hydrodynamical evolution is typically
incorporated by the use of a source term for the momentum and energy
equations. In a finite volume methodology, the momentum source term appears in 
integral form as
\begin{equation}
  \left.\frac{\partial (\rho \mathbf{u})}{\partial t}\right|_{\text{grav}} = \frac{1}{\Delta V} \int \rho \mathbf{g}\, dV
\end{equation}
and for the energy source term it is
\begin{equation}
  \left.\frac{\partial (\rho E)}{\partial t}\right|_{\text{grav}} = \frac{1}{\Delta V} \int \rho \mathbf{u}\cdot\mathbf{g}\, dV \label{eq:cell_center_gravity_source}.
\end{equation}
Here $\Delta V$ is the cell's volume.
In most hydrodynamics codes these are discretized as $\rho\,
\mathbf{g}$ and $\rho\, \mathbf{u}\,\cdot\mathbf{g}$, respectively, 
where $\rho$, $\mathbf{u}$, and $\mathbf{g}$ 
are evaluated at the zone
center.

There are two ways that this source term enters the hydrodynamics update.
First, during the hydrodynamics update, we alter the edge states that enter
into the determination of the fluxes. To second order in space and time, 
this can be done using the cell-centered
source term evaluated at time-level $n$. The details of this are described in 
the appendix. Second, after the hydrodynamics step, we add the time-centered source term
to the state. First we will describe how we do this for the momentum,
and then we will describe our approach for the energy. This discussion will be somewhat detailed.
However, we believe that the attention is necessary because of the importance of accuracy
in the gravitational source terms for our problem. The stability of the white dwarf binary
system is dependent in large part upon accurate coupling of the hydrodynamics and gravity;
an error in this approach could lead to, for example, a spurious mass transfer episode
that might lead us to very different conclusions about the long term stability of such a system.
Such considerations are generally unimportant for spherically-symmetric single star calculations,
but are of the utmost importance in a simulation where the global gravitational field can change 
quite significantly over the course of the simulation.

In a system with self-gravity, total momentum is conserved if the spatial domain
includes all of the mass of the system. This must be the 
case because each mass element exerts an equal and opposite gravitational force 
on every other mass element. However, the standard approach does not necessarily
guarantee that momentum is conserved numerically. We cannot represent a vacuum state 
in our code, so there is a small but non-zero density on the edge of the grid. 
This allows momentum to leak out of the domain even if the gravitational source term 
is written in an explicitly conservative manner. To see this, one can write the 
density in terms of the potential and then consider its spatial discretization. For simplicity,
we will consider one spatial dimension.
\begin{align}
  -\rho_{i}  \frac{d\Phi_{i}}{dx} &= -\frac{1}{4\pi G} \frac{d^2\Phi_i}{dx^2} \frac{d \Phi}{dx} \notag \\
  &= -\frac{1}{4\pi G} \left[\frac{\Phi_{i-1} - 2 \Phi_{i} + \Phi_{i+1}}{\Delta x^2}\right] \left[ \frac{\Phi_{i+1} - \Phi_{i-1}}{2\Delta x} \right] \notag \\
  &= -\frac{1}{8\pi G \Delta x^3} \left[ \Phi_{i+1}^2 - \Phi_{i-1}^2 - 2\Phi_i\left(\Phi_{i+1} - \Phi_{i-1}\right) \right] \label{eq:momentum_discretization}
\end{align}
It is easy to verify that adding the source terms two cells to the left
and right of this zone will result in complete cancellation of the source terms.
However, if the potential if non-zero outside of the domain, then there will be
momentum lost or gained from the grid, which will be encapsulated in the ghost cells
just outside the domain. In addition, when we replace the Laplacian above by the full
three-dimensional stencil including the $y$ and $z$ derivatives, and depending on the
discretization these will not be cancelled at all. This latter problem can be resolved by
writing the momentum update in an explicitly conservative way.

\citet[Chapter 4]{shu:1992} observes that it is possible to describe the source term 
for the momentum equation by taking the divergence of a gravitational stress tensor,
\begin{equation}
  G_{ij} = -\frac{1}{4\pi G}\left(g_i g_j - \frac{1}{2}|\mathbf{g}|^2\delta_{ij}\right).
\end{equation}
The momentum equations are then written explicitly in conservative form.
The flux at any zone boundary is added to one cell and
subtracted from another, so that the total momentum in the domain interior stays constant to
within numerical roundoff error. This result can be derived by analytically recasting 
Equation \ref{eq:momentum_discretization}. In the continuum limit, the two momentum
formulations are identical. Thus the latter has been advocated by, for example, 
\cite{jiang:2013} for the ATHENA code. However, in a finite discretization 
the divergence of the gravitational acceleration is no longer guaranteed to equal
the zone density. In particular, we find that the mixing of the gravitational components
means that the truncation error in the gravitational field can lead to large errors
that imply a density much different than the zone's actual density. This is especially
problematic in a simulation with a low-density ambient medium, where even a small error 
in the momentum update can lead to large changes in a zone's momentum. By continuing to explicitly
use the cell density in the momentum update, we can forgo this possibility: the size of the update
will always be suitably small if the zone's density is small. Thus for our simulations
we continue to use the standard source term for the momentum.

Time centering of this source term is done in CASTRO using a predictor-corrector approach.
At the start of a coarse grid timestep, we solve the gravitational potential for the density $\rho^n$.
We then add to the momenta a prediction of the source term that is first-order accurate in time, 
$\Delta t\, \rho^n\, \mathbf{g}^n$. After the hydrodynamics update, we recalculate
the gravitational potential based on the new density, $\rho^n$, and then add 
$-(\Delta t/2) \rho^n \mathbf{g}^n + (\Delta t/2) \rho^{n+1} \mathbf{g}^{n+1}$ to the momenta.

Dealing with the energy equation is more difficult. The central
challenge is to write down a form of the discretized energy equation
that explicitly conserves total energy when coupled to gravity. When
gravity is included, the total energy is
\begin{equation}
  \rho E_{\text{tot}} = \rho E - \frac{1}{2}\rho\Phi = \frac{1}{2}\rho \mathbf{u}^2 + \rho e - \frac{1}{2}\rho\Phi, \label{eq:total_energy_gravity}
\end{equation}
where $\rho E$ is the total gas energy from the pure hydrodynamics equation,
and the factor of $1/2$ in the gravitational energy term is necessary
for simulations with self-gravity to prevent double-counting of
interactions. The standard approach has been to note that the total energy update
can be written down by taking the dot product of the velocity and the momentum source term,
so that the time-centered energy update is
\begin{equation}
  \left.\Delta (\rho E)\right|_\text{grav} = \frac{\Delta t}{2} \left[ \rho^n \mathbf{u}^n \cdot \mathbf{g}^n + \rho^{n+1} \mathbf{u}^{n+1} \cdot \mathbf{g}^{n+1}\right].
\end{equation}
However, this does not conserve the total energy of Equation \ref{eq:total_energy_gravity}.
In particular, we notice in our simulations a long-term secular change in the gravitational potential energy
that is not accounted for in the gas energy. One solution to this problem has been offered by 
\cite{jiang:2013} and \cite{marcello:2012}. They propose to evolve the total energy instead of the gas energy 
in the hydrodynamics equations. When this is done, one can write down an explicitly conservative 
formulation:
\begin{equation}
  \frac{\partial \rho E_{\text{tot}}}{\partial t} = -\bm{\nabla}\cdot(\rho\mathbf{u}E + p\mathbf{u} + \mathbf{F_g}).
\end{equation}
Here $\mathbf{F_g}$ is a flux term which explicitly accounts for the change in the total energy due to self-gravity. 
By using Equation \ref{eq:cell_center_gravity_source} and the continuity equation, it is straightforward to show that
\begin{equation}
  {\bm\nabla} \cdot \mathbf{F_g} = -\rho \mathbf{u} \cdot \mathbf{g} - \frac{\partial}{\partial t}\left(\frac{1}{2}\rho\Phi\right). \label{eq:gradient_F_g}
\end{equation}
From here one can use the continuity equation to explicitly calculate the flux $\mathbf{F_g}$, as in \cite{jiang:2013}:
\begin{equation}
  \mathbf{F_g} = \frac{1}{8\pi G}\left(\Phi\bm{\nabla}\dot{\Phi} - \dot{\Phi}\bm{\nabla}\Phi\right) + \rho\mathbf{u}\Phi \label{eq:jiang_flux}.
\end{equation}
Here the dots refer to time derivatives. This has some attractive features. 
In particular, for a simulation with AMR, a flux-based approach
can quite naturally be accommodated in a way that conserves energy across levels, when fluxes
at interfaces between coarse and fine grids are matched. We note in passing that although
\citeauthor{jiang:2013} claim that this approach requires a Poisson solve for the time derivative of the potential,
$\dot{\Phi}$, this is not needed as in practice we will already have the potential at time-levels
$n$ and $n+1$, and to second order in time, $\dot{\Phi}^{n+1/2} = (\Phi^{n+1} - \Phi^{n}) / \Delta t.$

Now, the first term on the right-hand side of Equation \ref{eq:gradient_F_g} is the old source term. The second term simply 
accounts for the time rate of change of the gravitational potential energy. But since
\begin{equation}
  \frac{\partial(\rho E_{\text{tot}})}{\partial t} = \frac{\partial (\rho E)}{\partial t} + \frac{\partial (1/2) \rho \Phi}{\partial t},
\end{equation}
this does not actually add any information, since we can subtract off the time derivative 
of the potential energy and be back where we started. This demonstrates that the problem
with the original approach lies in the numerical discretization; if this is done correctly,
we can still properly conserve total energy.

Before pursuing that, we wish to point out the reason why we prefer not to use this 
approach of solving a total energy equation. The expression for the flux $\mathbf{F_g}$
given in \cite{jiang:2013} includes terms that are proportional to $\Phi$ and ${\bm{\nabla}} \Phi$,
but that are not explicitly proportional to the zone density $\rho$. Thus the approach suffers
from the same problems as the conservative momentum formulation discussed above. Truncation errors
in the gravitational field will result in an approximation to the density that can be quite 
different from the actual density, especially in low-density regions. In practice,
we have found this to be such a serious problem that this approach is not feasible for our problem.
The formulation of \cite{marcello:2012} does not suffer from the same problem. However, a 
remaining problem is that in this formalism, the internal energy
is calculated by subtracting the potential energy and the kinetic energy from the 
total energy. Due to numerical discretization errors, however, this can result in a serious 
error in the internal energy. This problem can be eased by the dual energy formalism,
but not completely. This means that we will potentially have very large errors in a 
zone's temperature, which can have serious repercussions in the reacting flows we will eventually consider,
as many reaction rates are very sensitive to temperature. Therefore our goal is instead to consider an approach
that is explicitly \textit{non}-conservative for the gas energy $(\rho E)$, but which
instead conserves the total energy $(\rho E_{\text{tot}})$ over the computational domain 
(subject, of course, to small errors due to energy leaking from the domain boundary).

The question of how to do this was resolved by \cite{arepo}. 
The essential problem with the standard approach is that 
fluid movements in a grid code are calculated using
fluxes at zone boundaries, while the standard gravity source term is
an estimate of the flux using the zone center quantities. This can
lead to a striking mismatch when the gradient across a zone is
large. We have found that this is especially important at the edge of
a star, where the large density gradient means that the momentum and
energy fluxes can transition sharply even across half of a zone
width. Another problem is that in a merger calculation in which two
stars are moving on a grid, large quantities of material are being
advected, so even a slight error in the calculation for the
high-density material can manifest itself as a long-term error in
energy conservation. This becomes apparent in Section
\ref{sec:kepler}.

We start with the integral formulation of the energy source term (Equation
\ref{eq:cell_center_gravity_source}), but do not discretize it at
the cell center.  Instead, we recognize that $\mathbf{u} = {\bm\nabla}\left[\mathbf{r}\cdot\mathbf{u}\right]$,
where $\mathbf{r}$ is the position vector originating from the zone center.
Then, since in the cell-centered approach we assume that the density, velocity,
and gravitational acceleration are constant across the zone, we can include all terms
inside the gradient operator. This allows us to use the divergence theorem to evaluate
the quantities on the zone faces. To be second-order accurate in time, the fields must 
all be evaluated at time-level $n+1/2$, but for clarity we will neglect the superscript.
\begin{align}
  \left.\Delta\left(\rho E\right)_i\right|_{\text{grav}} &= \frac{\Delta t}{\Delta V} \int \rho \mathbf{u}\cdot\mathbf{g}\, dV_i \notag \\
     &= \frac{\Delta t}{\Delta V} \int \rho\, {\bm \nabla} \left[\mathbf{r} \cdot \mathbf{u}\right]\cdot\mathbf{g}\, dV_i \notag \\
     &= \frac{\Delta t}{\Delta V} \int {\bm \nabla} \cdot \left[ \rho\, (\mathbf{r} \cdot \mathbf{u})\, \mathbf{g}\right]\, dV_i \notag \\
     &= \frac{\Delta t}{\Delta V} \int \rho\, (\mathbf{r} \cdot \mathbf{u})\, \mathbf{g} \cdot \mathbf{dS}_i.
\end{align}
Here $\mathbf{dS}$ is a surface element whose vector points normal to
and away from the surface. In one dimension, this expression evaluates to
\begin{align}
  \left.\Delta\left(\rho E\right)_i\right|_{\text{grav}} &= \frac{\Delta t}{\Delta V} \rho_{i-1/2}\, \frac{\Delta x}{2} u_{x_{i-1/2}} g_{x_{i-1/2}} \Delta A_{i-1/2} \notag \\
             &+ \frac{\Delta t}{\Delta V} \rho_{i+1/2}\, \frac{\Delta x}{2} u_{x_{i+1/2}} g_{x_{i+1/2}} \Delta A_{i+1/2} \notag \\
             &= \frac{1}{2} \rho_{i-1/2} u_{x_{i-1/2}} g_{x_{i-1/2}} \notag \\
             &+ \frac{1}{2} \rho_{i+1/2} u_{x_{i+1/2}} g_{x_{i+1/2}}.\label{eq:edge_centered_gravity_source}
\end{align}
(We note that this differs slightly from Equation 94 in \cite{arepo} in that
we evaluate the gravitational acceleration at the zone face rather than the zone center.)
In the limit of constant fields, this expression simply averages out to the standard 
source term at the zone center. However, when the fields vary this can give very different results.

As pointed out by \citeauthor{arepo}, $(\rho u)$ evaluated at the faces at time level $n+1/2$ is the mass flux,
so $(\rho u)_{i-1/2}\, \Delta t\, \Delta A = \Delta m_{i-1/2}$ is the mass exchanged between zones $i$ and $i-1$.
Additionally, to second order $g_{x_{i-1/2}} \Delta x = \Phi_{i} - \Phi_{i-1}$. Therefore the energy gained by the zone is
\begin{align}
  \left.\Delta\left(\rho E\right)_i\right|_{\text{grav}} &= \frac{1}{2} \Delta m_{i-1/2} \left(\Phi_{i} - \Phi_{i-1}\right) \notag \\
     &+ \frac{1}{2} \Delta m_{i+1/2} \left(\Phi_{i+1} - \Phi_{i}\right)
\end{align}
This expression says that the energy gained by a zone is due to a mass element moving through a potential gradient
between two zones. Now, the key observation is that this does not have the telescoping property of a conservative flux;
instead, the \textit{same} amount of gravitational potential energy is gained by zones $i$ and $i-1$ by the mass flux 
between them. When this source term is summed up over all zones, it will exactly compensate for the change in 
gravitational potential energy, except for any energy lost through the boundaries. The extension of this energy change 
to multiple dimensions simply means that we include the mass fluxes and potential gradients across the $y$ and $z$ faces.
We have found that in practice this method for conserving the total energy is the least susceptible to deleterious effects
from numerical discretization errors. However, it does mean that there can be sharp changes in the total energy of a zone
in regions near a white dwarf's surface, because the mass fluxes into the ambient medium can be quite large compared to
the mass of the zone. This effect diminishes at higher resolutions as the gradients become better resolved.

@@

Since the results of a merger simulation depend strongly on gravity,
it is important to check whether proper numerical convergence is
achieved for the Poisson solver. To do so, we created a simple test
that inserts a sphere of radius $R$ and uniform mass density $\rho$
onto our grid, and used CASTRO to calculate the gravitational
potential $\Phi$ of this setup. We ensure that $R$ is an integer
multiple of the grid spacing, and the center of the sphere is at the
origin. The problem domain for our simulations is $[-1.6, 1.6]^3$, and
we take $R = 0.5$ and $\rho = 10^3$. The zones with $r > R$ are filled
with an ambient material of very low density. We run this problem at
multiple resolutions corresponding to jumps by a factor of two. For
comparison, at each grid point we evaluate the analytical potential of
a uniform sphere, which can be easily determined using Gauss' law:
\begin{equation}
  \Phi_{\text{sphere}}(r) = \frac{GM}{r} \times \begin{cases} (3R^2 - r^2)/(2 r^2) & r \leq R \\ 1 & r > R \end{cases},\label{eq:sphere-analytical}
\end{equation}
where $M = 4\pi R^3 / 3$ is the mass of the sphere. Defining the $L^2$
norm of a field $f$ as
\begin{equation}
  \| f \|_2 = \left(\sum_{i,j,k} dx\, dy\, dz\, f_{ijk}^2\right)^{1/2},
\end{equation}
we measure the numerical error by calculating the $L^2$ norm of the error and normalizing it by the $L^2$ norm of the analytical solution:
\begin{equation}
  \text{Error} = \frac{\|\Phi - \Phi_{\text{sphere}}\|_2}{\|\Phi_{\text{sphere}}\|_2}.
\end{equation}

@@

The only way to fully eliminate this effect is to test a problem that
does not change with resolution. The obvious companion is a cube of
uniform density $\rho$, where now $R$ is half of the side length of
the cube. At each resolution we use the same $R$ as for the sphere,
which ensures that the cube always fills exactly the same fraction of
the domain and thus has the same mass, so the only improvement comes
from better sampling at higher resolution. The potential for this
object has been worked out analytically by \citet{waldvogel:1976} (see
also a similar result in \citealt{hummer:1996}). The potential is in
Equation 15 of that paper, though the last term is missing a factor of
$1/2$, which destroys the symmetry. Inserting this missing factor and
performing a simple coordinate transformation so that the center of
the cube is at the origin, the potential is
\begin{align}
  \Phi_{\text{cube}}(x,y,z) &= G\rho\sum_{i,j,k=0}^1\left[x_i y_j\, \text{tanh}^{-1}\left(\frac{z_k}{r_{ijk}}\right)\right. \notag \\
  &+ \left. y_j z_k\, \text{tanh}^{-1}\left(\frac{x_i}{r_{ijk}}\right) + z_k x_i\, \text{tanh}^{-1}\left(\frac{y_j}{r_{ijk}}\right) \right.\notag \\
  &\left. - \frac{x_i^2}{2}\,\text{tan}^{-1}\left(\frac{y_j z_k}{x_i r_{ijk}}\right) - \frac{y_j^2}{2}\,\text{tan}^{-1}\left(\frac{z_k x_i}{y_j r_{ijk}}\right) \right. \notag \\
  &- \left. \frac{z_k^2}{2}\,\text{tan}^{-1}\left(\frac{x_i y_j}{z_k r_{ijk}}\right)\right]
\end{align}
where $x_0 = R + x$, $x_1 = R - x$, $y_0 = R + y$, $y_1 = R-y$, $z_0 =
R+z$, $z_1 = R-z$, and $r_{ijk} = \sqrt{x_i^2 + y_j^2 + z_k^2}$. In
\texttt{FORTRAN} and \texttt{C}, the inverse hyperbolic tangent is
\texttt{atanh} and the inverse tangent is \texttt{atan} (\textit{not}
\texttt{atan2}). This formula is valid both inside and outside the
cube. The normalized $L^2$ error norm for this problem is also shown
in Figure \ref{fig:gravity_convergence}, and indeed we see
near-perfect second-order scaling.

@@

The tracing of gravity can be controlled in the public version of Castro
with the parameter {\tt castro.ppm\_trace\_grav}.
